---
title: "Assignment 2"
author: Valentin, Anda, Lea
date: "2022-12-04"
output: revealjs::revealjs_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Package to install:
*install.packages("plotly")


## Introduction to Plotly |
#### 1. What is Plotly?
* Plotly is an open-source data visualization library for Python and R written in JavaScript to create interactive graphs (Sanchez-Ayala, 2020)
*

## Gapminder Data Set

```{r echo=FALSE, message=FALSE, comment=NA}
library(gapminder)
library(dplyr)
head(gapminder)
```
## Using plotly.js with ggplot integration
With the function ggplotly() we can easily convert a graph created with the ggplot2 package into interactive figures on plotly.js.After we created our plot with ggplot , let's call it 'p' - we simply have to assign it to a new figure with the function: fig <- ggplotly(p) - and we got our interactive figure.  
To make it more interesting, we chose a little more complex plot including geom_point() that includes differently sized bubbles, and an annotation of country codes by using geom_text(). These are also called "Bubblecloud Plots". 

## Bubble Cloud Plot from gapminder
```{r, echo=FALSE, warning=FALSE, message=FALSE, comment=NA}
# Loading libraries

library(countrycode)
library(tidyverse)
library(gapminder)
library(dplyr)

gapminder <- gapminder %>% filter(year==2007)
gapminder <- gapminder %>%   # We select some relevant columns
select(country, year, lifeExp, pop, continent, gdpPercap) %>%
# We use mutate() to transform the variable `code`
mutate(
# We use countrycode() to extract the iso3 code from the english name
country_code = countrycode(country,
origin = "country.name",
destination = "iso3c"))

gapminder$continent_code <- recode(gapminder$continent,
"Asia" = "ASI",
"Africa" = "AFR",
"Americas" = "AME",
"Oceania"= "OCE",
"Europe"= "EUR")  

gapminder_eu <- gapminder %>% filter(continent=="Europe")
gapminder_eu_countries <- distinct(gapminder_eu, country)

gapminder_eu$subregion <- recode(gapminder_eu$country,
"Albania" = "Eastern Europe", 
"Austria" = "Western Europe",
"Belgium" = "Western Europe",
"Bosnia and Herzegovina" = "Eastern Europe", 
"Bulgaria" = "Eastern Europe", 
"Croatia" = "Eastern Europe", 
"Czech Republic" = "Eastern Europe", 
"Denmark" = "Northern Europe",
"Finland" = "Northern Europe",
"France" = "Western Europe",
"Germany" = "Western Europe",
"Greece" = "Southern Europe",
"Hungary" = "Eastern Europe", 
"Iceland" = "Northern Europe",
"Ireland" = "Western Europe",
"Italy" = "Southern Europe",
"Montenegro" = "Eastern Europe",
"Netherlands" = "Western Europe",
"Norway" = "Northern Europe",
"Poland" = "Eastern Europe", 
"Portugal" = "Southern Europe", 
"Romania" = "Eastern Europe", 
"Serbia" = "Eastern Europe", 
"Slovak Republic" = "Eastern Europe", 
"Slovenia" = "Eastern Europe", 
"Spain" = "Southern Europe",
"Sweden" = "Northern Europe",
"Switzerland" = "Western Europe",
"Turkey" = "Eurasia", 
"United Kingdom" = "Western Europe")


library(plotly)
library(gapminder)

gapminder_eu$pop <- gapminder_eu$pop/1000000

p1 <- gapminder_eu %>% 
       ggplot(aes(x=pop, y=lifeExp)) + 
  geom_point(aes(size=pop, color=subregion, text=paste("country:", country))) +
geom_text(aes(size=pop, label=country_code, show.legend=FALSE), colour="gray20") +
  labs(title = "Life Expectancy & Population in Europe",
       caption = "*different point sizes according to Population Size",
       x = "Population in Mio.",
       y = "Average Life Expectancy",
       size ="") +
  scale_colour_discrete(name="Sub Region*")


fig <- ggplotly(p1) 

fig

 ggsave("fig-1.png", width = 50, height = 40, units = "cm")
  
 
```


## Bubble Cloud Plot Animation 
We can also include animations in our Bubblecloud Plots, by simply adding the frame = "" specification to our geom_text function. Here we want to see development of Population Size and GDP in across continents in different years, so we use frame = year.

## Plot Animation
```{r, echo=FALSE, warning=FALSE, message=FALSE, comment=NA}

library(tidyverse)
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_cont <- gapminder %>% filter(year %in% c(1987, 1992, 1997, 2007, 2002))
 
gapminder_cont$continent_code <- recode(gapminder_cont$continent,
"Asia" = "ASI",
"Africa" = "AFR",
"Americas" = "AME",
"Oceania"= "OCE",
"Europe"= "EUR")  

gapminder_cont <- gapminder_cont %>%    #create a newdata table of the aggregated data
           group_by(continent, continent_code, year) %>%    
           summarize(gdpPercap = mean(gdpPercap),
                     pop = mean(pop))

gapminder_cont$pop <- gapminder_cont$pop/1000000
p2 <- gapminder_cont %>% 
       ggplot(aes(x=pop, y=gdpPercap, frame = year)) + 
  geom_point(aes(size=pop, frame = year, color=continent)) +
geom_text(aes(size=pop, frame = year,label=continent_code), colour="gray20") +
  labs(title = "Population & GDP across continents from 1987-2007",
       caption= "*different point sizes according to Population Size",
       x = "Population in Mio.",
       y = "GDP per capita",
       size="")+
  scale_colour_discrete(name="Region")

fig2 <- ggplotly(p2)

fig2


```

---

## References

* Sanchez-Ayala, M., Medium, 16.02.2020: https://medium.com/swlh/plotly-beautiful-data-visualization-made-easy-3f7e48864706

## Using plotly.js with ggplot integration
With the function ggplotly() we can easily convert a graph created with the ggplot2 package into interactive figures on plotly.js.After we created our plot with ggplot , let's call it 'p' - we simply have to assign it to a new figure with the function: fig <- ggplotly(p) - and we got our interactive figure.  
To make it more interesting, we chose a little more complex plot including geom_point() that includes differently sized bubbles, and an annotation of country codes by using geom_text(). These are also called "Bubblecloud Plots". 

## Bubble Cloud Plot from gapminder
```{r, echo=FALSE, warning=FALSE, message=FALSE, comment=NA}
# Loading libraries

library(countrycode)
library(tidyverse)
library(gapminder)
library(dplyr)

gapminder <- gapminder %>% filter(year==2007)
gapminder <- gapminder %>%   # We select some relevant columns
select(country, year, lifeExp, pop, continent, gdpPercap) %>%
# We use mutate() to transform the variable `code`
mutate(
# We use countrycode() to extract the iso3 code from the english name
country_code = countrycode(country,
origin = "country.name",
destination = "iso3c"))

gapminder$continent_code <- recode(gapminder$continent,
"Asia" = "ASI",
"Africa" = "AFR",
"Americas" = "AME",
"Oceania"= "OCE",
"Europe"= "EUR")  

gapminder_eu <- gapminder %>% filter(continent=="Europe")
gapminder_eu_countries <- distinct(gapminder_eu, country)

gapminder_eu$subregion <- recode(gapminder_eu$country,
"Albania" = "Eastern Europe", 
"Austria" = "Western Europe",
"Belgium" = "Western Europe",
"Bosnia and Herzegovina" = "Eastern Europe", 
"Bulgaria" = "Eastern Europe", 
"Croatia" = "Eastern Europe", 
"Czech Republic" = "Eastern Europe", 
"Denmark" = "Northern Europe",
"Finland" = "Northern Europe",
"France" = "Western Europe",
"Germany" = "Western Europe",
"Greece" = "Southern Europe",
"Hungary" = "Eastern Europe", 
"Iceland" = "Northern Europe",
"Ireland" = "Western Europe",
"Italy" = "Southern Europe",
"Montenegro" = "Eastern Europe",
"Netherlands" = "Western Europe",
"Norway" = "Northern Europe",
"Poland" = "Eastern Europe", 
"Portugal" = "Southern Europe", 
"Romania" = "Eastern Europe", 
"Serbia" = "Eastern Europe", 
"Slovak Republic" = "Eastern Europe", 
"Slovenia" = "Eastern Europe", 
"Spain" = "Southern Europe",
"Sweden" = "Northern Europe",
"Switzerland" = "Western Europe",
"Turkey" = "Eurasia", 
"United Kingdom" = "Western Europe")


library(plotly)
library(gapminder)

gapminder_eu$pop <- gapminder_eu$pop/1000000

p1 <- gapminder_eu %>% 
       ggplot(aes(x=pop, y=lifeExp)) + 
  geom_point(aes(size=pop, color=subregion, text=paste("country:", country))) +
geom_text(aes(size=pop, label=country_code, show.legend=FALSE), colour="gray20") +
  labs(title = "Life Expectancy & Population in Europe",
       caption = "*different point sizes according to Population Size",
       x = "Population in Mio.",
       y = "Average Life Expectancy",
       size ="") +
  scale_colour_discrete(name="Sub Region*")


fig <- ggplotly(p1) 

fig

 ggsave("fig-1.png", width = 50, height = 40, units = "cm")
  
 
```


## Bubble Cloud Plot Animation 
We can also include animations in our Bubblecloud Plots, by simply adding the frame = "" specification to our geom_text function. Here we want to see development of Population Size and GDP in across continents in different years, so we use frame = year.

## Plot Animation
```{r echo=FALSE, warning=FALSE,message=FALSE, comment=NA}

library(tidyverse)
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder_cont <- gapminder %>% filter(year %in% c(1987, 1992, 1997, 2007, 2002))
 
gapminder_cont$continent_code <- recode(gapminder_cont$continent,
"Asia" = "ASI",
"Africa" = "AFR",
"Americas" = "AME",
"Oceania"= "OCE",
"Europe"= "EUR")  

gapminder_cont <- gapminder_cont %>%    #create a newdata table of the aggregated data
           group_by(continent, continent_code, year) %>%    
           summarize(gdpPercap = mean(gdpPercap),
                     pop = mean(pop))

gapminder_cont$pop <- gapminder_cont$pop/1000000
p2 <- gapminder_cont %>% 
       ggplot(aes(x=pop, y=gdpPercap)) + 
  geom_point(aes(size=pop, color=continent,frame = year, text=paste("continent:", continent))) +
geom_text(aes(size=pop, label=continent_code,frame = year), colour="gray20") +
  labs(title = "Population & GDP across continents from 1987-2007",
       subtitle = "Different point sizes according to Population Size",
       x = "Population in Mio.",
       y = "GDP per capita",
       size="")+
  scale_colour_discrete(name="Region")

fig2 <- ggplotly(p2, width = 600, height = 400) %>%
    animation_opts(1000)

fig2


```

---

## References

* Sanchez-Ayala, M., Medium, 16.02.2020: https://medium.com/swlh/plotly-beautiful-data-visualization-made-easy-3f7e48864706
